Area maximisation in continuous time

### 2012-11-12 Notes/ContTime.tex
started cosyne abstract

\begin{thm}
  The solution of \eqref{eq:fptrecdg} is unique.
\end{thm}
\begin{proof}
  We have already seen that the diagonal part is determined by \eqref{eq:fptrecdg}.
  Suppose we have two solutions for the off-diagonal part, $\fptb$ and $\fptb'$.
  Subtracting \eqref{eq:fptrecdg} for these two solutions gives
  %
  \begin{equation*}
    \Lb\MM(\fptb-\fptb')=0.
  \end{equation*}
  %
  As we are assuming an ergodic process, the zero eigenvalue is non-degenerate.
  Therefore
  %
  \begin{equation*}
    \fptb-\fptb' = \onev\mathbf{y},
  \end{equation*}
  %
  for some row-vector $\mathbf{y}$. Looking at the diagonal components
  %
  \begin{equation*}
    \mathbf{y}_{i} = \fptb_{ii}-\fptb'_{ii} = 0,
  \end{equation*}
  %
  and therefore $\fptb-\fptb'=0$.
\end{proof}

\begin{thm}
  The off-diagonal mean first passage times are given by
  %
  \begin{equation}\label{eq:fptfund}
    \fptb = (\onem\fund\dg - \fund)\D.
  \end{equation}
  %
\end{thm}
\begin{proof}
  We just need to show that this satisfies \eqref{eq:fptrecdg} with \eqref{eq:recurtime}, and that the diagonal elements vanish.
  Premultiply \eqref{eq:fptfund} by $\Lb\MM$ and subtract from \eqref{eq:fptrecdg}:
  %
  \begin{equation*}
  \begin{aligned}
    \fpt\dg - \Lb\MM\fptb &= \Lb(\I+\MM\fund)\D\\
     &= \Lb\onev\eq\D \\
     &= \Lb\onem,
  \end{aligned}
  \end{equation*}
  %
  where we used \eqref{eq:fundQZ} and \eqref{eq:pdotD}.
  The vanishing of the diagonal elements follows trivially from the component form of \eqref{eq:fptfund}:
%
\begin{equation}\label{eq:fptfundcmpt}
  \fptb_{ij} = \frac{\fund_{jj}-\fund_{ij}}{\eq_j}.
\end{equation}
%

@@

Degeneracy tends to be very delicate. It is usually hard to arrange without some symmetry relating degenerate state. Such a symmetry would imply lumpability (see \sref{sec:lump}). The lumped chain would not have any shortcuts, as an entire degenerate set cannot be bypassed. As this lumped chain has the same area (see \sref{sec:SNRlump}), we would need only consider the multi-state topology.

@@

The multi-state topology is defined by:
%
\begin{equation}\label{eq:multistatedef}
  \W^+_{ij} = q^+_i \delta_{i+1,j},
  \qquad
  \W^-_{ij} = q^-_j \delta_{i,j+1}.
\end{equation}
%
It saturates various inequalities:
%
\begin{equation}\label{eq:multistateineq}
  \begin{aligned}
    \fptb_{ik} - \fptb_{jk} &=
      \begin{cases}
        \fptb_{ij},  &\text{if}\quad i \leq j \leq k \quad\text{or}\quad i \geq j \geq k,\\
        -\fptb_{ji}, &\text{if}\quad j \leq i \leq k \quad\text{or}\quad j \geq i \geq k,\\
      \end{cases} \\
    r\eq_i \W^F_{ij} \prn{\fptb_{ij}+\fptb_{ji}} &= 1 \quad\text{if}\quad i=j\pm1,
  \end{aligned}
\end{equation}
%
and it satisfies detailed balance (a.k.a.\ reversibility a.k.a. $\CL^2_{\eq}$ self-adjointness):
%
\begin{equation}\label{eq:multistateprob}
  f^+ q^+_i \eq_i = f^- q^-_i \eq_{i+1},
\end{equation}
%
which means we can always choose the transition rates, $q^\pm_i$, to give any desired equilibrium probabilities, $\eq_i$.

@@

First let us maximise \eqref{eq:multistatearea} at fixed $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
Clearly this will happen when we put all of the probability at the ends: $\eq_1=\eq_-$ and $\eq_n=\eq_+$ are the only non-zero $\eq_k$.
This gives an area of
\footnote{Note that including the dropped factors in \eqref{eq:noisepm} would only change this to $A \leq \sqrt{N(4\eq_1\eq_n)}(n-1)/r$, which would have no effect on what follows.}

%
\begin{equation}\label{eq:multistateextarea}
  A \leq \frac{\sqrt{N}}{r}(n-1)\prn{4\eq_1\eq_n}.
\end{equation}
%
This is maximised at $\eq_1=\eq_n=\half$:
%
\begin{equation}\label{eq:maxarea}
  A \leq \frac{\sqrt{N}}{r}(n-1).
\end{equation}
%
Note that this is not an ergodic chain, the two states at each end are absorbing.

@@

\bibliographystyle{utcaps_sl}
\bibliography{maths}

>>>>

We can substitute this into \eqref{eq:fptrecdg} to get a recursion relation for the off diagonal part:
%
\begin{equation}\label{eq:fptbrec}
  \MM\fptb = \D-\onem.
\end{equation}
%
\begin{thm}
  The off-diagonal solution of \eqref{eq:fptbrec} is unique.
\end{thm}
\begin{proof}
  Suppose we have two solutions for the off-diagonal part, $\fptb$ and $\fptb'$.
  Subtracting \eqref{eq:fptbrec} for these two solutions gives
  %
  \begin{equation*}
    \MM(\fptb-\fptb')=0.
  \end{equation*}
  %
  As we are assuming an ergodic process, the zero eigenvalue is non-degenerate.
  Therefore
  %
  \begin{equation*}
    \fptb-\fptb' = \onev\mathbf{y},
  \end{equation*}
  %
  for some row-vector $\mathbf{y}$. Looking at the diagonal components
  %
  \begin{equation*}
    \mathbf{y}_{i} = \fptb_{ii}-\fptb'_{ii} = 0,
  \end{equation*}
  %
  and therefore $\fptb-\fptb'=0$.
\end{proof}

\begin{thm}
  The off-diagonal mean first passage times are given by
  %
  \begin{equation}\label{eq:fptfund}
    \fptb = (\onem\fund\dg - \fund)\D.
  \end{equation}
  %
\end{thm}
\begin{proof}
  We just need to show that this satisfies \eqref{eq:fptbrec} and that the diagonal elements vanish.
  Premultiply \eqref{eq:fptfund} by $\MM$:
  %
  \begin{equation*}
  \begin{aligned}
    \MM\fptb &= -\MM\fund\D\\
     &= (\I-\onev\eq)\D \\
     &= \D-\onem,
  \end{aligned}
  \end{equation*}
  %
  where we used \eqref{eq:fundQZ} and \eqref{eq:pdotD}.
  The vanishing of the diagonal elements follows trivially from the component form of \eqref{eq:fptfund}:
%
\begin{equation}\label{eq:fptfundcmpt}
  \fptb_{ij} = \frac{\fund_{jj}-\fund_{ij}}{\eq_j}.
\end{equation}
%

@@

Degeneracy tends to be very delicate. It is usually hard to arrange without some symmetry relating degenerate states. Such a symmetry would imply lumpability (see \sref{sec:lump}). The lumped chain would not have any shortcuts, as an entire degenerate set cannot be bypassed. As this lumped chain has the same area (see \sref{sec:SNRlump}), we would need only consider the multi-state topology.

@@

The multi-state topology is defined by (see \cite{amit1994learning,Fusi2007multistate}):
%
\begin{equation}\label{eq:multistatedef}
  \W^+_{ij} = q^+_i \delta_{i+1,j},
  \qquad
  \W^-_{ij} = q^-_j \delta_{i,j+1}.
\end{equation}
%
It saturates various inequalities:
%
\begin{equation}\label{eq:multistateineq}
  \begin{aligned}
    \fptb_{ik} - \fptb_{jk} &=
      \begin{cases}
        \fptb_{ij},  &\text{if}\quad i \leq j \leq k \quad\text{or}\quad i \geq j \geq k,\\
        -\fptb_{ji}, &\text{if}\quad j \leq i \leq k \quad\text{or}\quad j \geq i \geq k,\\
      \end{cases} \\
    r\eq_i \W^F_{ij} \prn{\fptb_{ij}+\fptb_{ji}} &= 1 \quad\text{if}\quad i=j\pm1,
  \end{aligned}
\end{equation}
%
and it satisfies detailed balance (a.k.a.\ reversibility a.k.a. $\CL^2_{\eq}$ self-adjointness):
%
\begin{equation}\label{eq:multistateprob}
  f^+ q^+_i \eq_i = f^- q^-_i \eq_{i+1},
\end{equation}
%
which means we can always choose the transition rates, $q^\pm_i$, to give any desired equilibrium probabilities, $\eq_i$.

@@

First let us maximise \eqref{eq:multistatearea} at fixed $\eq_\pm = \sum_k \eq_k \prn{\frac{1\pm\w_k}{2}}$.
Clearly this will happen when we put all of the probability at the ends: $\eq_1=\eq_-$ and $\eq_n=\eq_+$ are the only non-zero $\eq_k$.
This gives an area of
\footnote{Note that including the dropped factors in \eqref{eq:noisepm} would only change this to $A \leq \sqrt{N(4\eq_+\eq_-)}(n-1)/r$, which would have no effect on what follows.}

%
\begin{equation}\label{eq:multistateextarea}
  A \leq \frac{\sqrt{N}}{r}(n-1)\prn{4\eq_+\eq_-}.
\end{equation}
%
This is maximised at $\eq_+=\eq_-=\half$:
%
\begin{equation}\label{eq:maxarea}
  A \leq \frac{\sqrt{N}}{r}(n-1).
\end{equation}
%
Note that the chain that achieves this is not ergodic, the two states at each end are absorbing. This is similar to the results found numerically in \cite{Barrett2008discrete} in a slightly different situation.

@@

\bibliographystyle{utcaps_sl}
\bibliography{maths,neuro}
